Staged Abstract Interpreters

### 2018-10-29 paper/10_intro.tex
worked on paper

Multi-stage programming has been widely used to improve the performance in many domains, 
such as optimizing compilers and domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13, DBLP:journals/jfp/CaretteKS09},
numerical computation \cite{PGL-038, DBLP:conf/pepm/AktemurKKS13}, 
generic programming \cite{DBLP:journals/pacmpl/Yallop17}, 
data processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14, DBLP:conf/popl/KiselyovBPS17}, 
query compilation in databases \cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
and etc.
Likewise, specializing static analyses by partial evaluation had been emerged in 90s 
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical}, 
and indeed it is able to effectively remove the interpretive overhead of 
repeatedly traversing the abstract syntax tree. However, it is still worth 
to investigate the idea deeper from a modern perspective based on meta-programming 
for general abstract interpreters.
Practically, precise whole-program analyses formulated as abstract interpretation 
can be very slow due to the inherent complexity of the algorithm and the large scale 
of the analyzed programs \cite{toman_et_al:LIPIcs:2017:7121} \todo{other cite?};
particularly, we would like a systematic approach to optimize program 
analyzers and meanwhile minimally modifying the analyzer programs.
In the present work, we show how to achieve this goal by bring existing 
high-performance computing facilities, in the style of multi-stage generative
programming, to abstract interpretation semantic artifacts, i.e., 
abstract interpreters.

The slogan of multi-stage programming is "abstraction without regret". 
Of course, all interpreters are metalinguistic abstractions, but some interpreters are more abstract 
than others \todo{maybe rephrase}. 
One technical challenge is to deal with the non-determinisism, fixed-point iterations and abstract 
domains in abstract interpreters when staging is introduced. The abstract interpreter, as a semantic artifact, 
should be written in a style that is easy for people to communicate the formulation, but also be friendly
to specialization and generating efficient code. \todo{We discuss the form of abstract interpreters for staging}
We show the use LMS framework, that enables multi-stage programming by just changing the types 
of the analyzer program, and we thus bridge the gap between designing sound program analyses 
and implementing efficient program analyzers.

@@

\begin{itemize}
\item Starting from a definitional abstract interpreter for a higher-order functional language,
  which the syntax and semantics are described in section \ref{bg_lang},
  we show how to turn it into a staged abstract interpreter by adding stage annotations within
  Lightweight Modular Staging (LMS) framework, a library-based approach to code generation. 
  We later extend the staged abstract interpreter to support imperative language features 
  (section \ref{cases_imp}), such as states, loops and exceptions.
\item We elaborate the idea that the staged abstract interpreter not only improves the \textit{performance}
  but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
  programs. \todo{more on this point.}
\item To evaluate the generality and applicability of the idea, we conduct three case studies in section \ref{cases_study}. 
  First we revisited Abstract Compilation \cite{Boucher:1996:ACN:647473.727587}, which is a closure generation 
  technique applied on 0-CFA. We show that it can be understood and easily implemented as an instance of 
  staged abstract interpreter. 
  Second, we demonstrate sophisticated control-flow analysis techniques, such as adaptive context/flow/path-sensitive 
  analysis and store-widening can be integrated with staged abstract interpreters without losing any precision.
  Finally, we show that in an imperative language, the idea can be applied onto an abstract interpreter with
  an abstract domain that performs numerical analysis.
  (Optional: taint/information flow analysis?)
\item We also empirically evaluate the speed improved by staging an big-step abstract interpreter 
  for control-flow analysis. We compare against with the unstaged version, a small-step version (unstaged), 
  and an optimizing small-step version.
\end{itemize}

>>>>

Multi-stage programming has been widely used to improve the performance in many domains, 
such as optimizing compilers and domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13, DBLP:journals/jfp/CaretteKS09},
numerical computation \cite{PGL-038, DBLP:conf/pepm/AktemurKKS13}, 
generic programming \cite{DBLP:journals/pacmpl/Yallop17}, 
data processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14, DBLP:conf/popl/KiselyovBPS17}, 
query compilation in databases \cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
and etc.
Likewise, specializing static analyses by partial evaluation had been emerged in 1990s 
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical}, 
and indeed it is able to effectively remove the interpretive overhead of 
repeatedly traversing the abstract syntax tree. However, it is still worth 
to investigate the idea deeper from a modern perspective based on meta-programming 
for general abstract interpreters.
Practically, precise whole-program analyses formulated as abstract interpretation 
can be very slow due to the inherent complexity of the algorithm and the large scale 
of the analyzed programs \cite{toman_et_al:LIPIcs:2017:7121} \todo{other cite?};
particularly, we would like a systematic approach to optimize program 
analyzers and meanwhile minimally modifying the analyzer programs.
In the present work, we show how to achieve this goal by bring existing 
high-performance computing facilities, in the style of multi-stage generative
programming, to abstract interpretation semantic artifacts, i.e., 
abstract interpreters.

The slogan of multi-stage programming is "abstraction without regret". 
Of course, all interpreters are metalinguistic abstractions, but some interpreters are more "abstract"
than others \todo{maybe rephrase}. 
One technical challenge is to deal with the non-determinisism, fixed-point iterations and abstract 
domains in abstract interpreters when staging is introduced. The abstract interpreter, as a semantic artifact, 
should be written in a style that is easy for people to communicate the formulation, but also be friendly
to specialization and generating efficient code. \todo{We discuss the form of abstract interpreters for staging}
We show the use LMS framework, that enables multi-stage programming by just changing the types 
of the analyzer program, and we thus bridge the gap between designing sound program analyses 
and implementing efficient program analyzers.

@@

\begin{itemize}
\item Starting from a definitional abstract interpreter for a higher-order functional language,
  which the syntax and semantics are described in section \ref{bg_lang},
  we show how to turn it into a staged abstract interpreter by adding stage annotations within
  Lightweight Modular Staging (LMS) framework, a library-based approach to code generation. 
  We later extend the staged abstract interpreter to support imperative language features 
  (section \ref{cases_imp}), such as states, loops and exceptions.
\item We elaborate the idea that the staged abstract interpreter not only improves the \textit{efficiency}
  but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
  programs. \todo{more on this point.}
\item To demonstrate the feasibility and applicability of the proposed approach, we conduct three case studies in section \ref{cases_study}. 
  First we revisited Abstract Compilation \cite{Boucher:1996:ACN:647473.727587}, which is a closure generation 
  technique applied on 0-CFA. We show that it can be understood and easily implemented as an instance of 
  staged abstract interpreter. 
  Second, we demonstrate sophisticated control-flow analysis techniques, such as adaptive context/flow/path-sensitive 
  analysis and store-widening can be integrated with staged abstract interpreters without losing any precision.
  Finally, we show that in an imperative language, the idea can be applied onto an abstract interpreter with
  an abstract domain that performs numerical analysis.
  (Optional: taint/information flow analysis?)
\item We also empirically evaluate the speed improved by staging an big-step abstract interpreter 
  for control-flow analysis. We compare against with the unstaged version, a small-step version (unstaged), 
  and an optimizing small-step version.
\end{itemize}

### 2018-10-29 paper/50_cases.tex
worked on paper

\citeauthor{Boucher:1996:ACN:647473.727587} introduced abstract compilation as a new
implementation technique for abstract interpretation based static analysis \cite{Boucher:1996:ACN:647473.727587}.
Their work is esentially a partial evaluated CE (\textit{Control-Environment}) abstract
interpreter written in a big-step semantics for programs written in continuation-passing style.
The analysis it can perform is a monovariant control flow analysis, sepcifically, Shiver's original
formulation of 0-CFA \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
After the partial evaluation on input program, the \textit{control} part of abstract interpreter
is disappeared, instead, the compiled analysis is a closure (higher-order function) and
only takes an environment as argument,
and the overhead of traversing the abstract syntax tree of source program also has been eliminated.
Since the target language is written a continuation-passing style, so there is
no explict evaluation context component when constructing the abstract interpreter
(rather than a CEK machine).

In this section, we firstly review their work of closure generation,
and reproduce their result by adding stage annotations to 0-CFA.
Show that \citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation can be understood as an instance of staged 
abstract interpreter.

Advantages: (almost?) no need to modify the analysis code; staging the store data structure

(Fianally, we should briefly summarize the difference between them:
staging vs partial evaluation; code generation; engineering effort; performance)

(Specific -> General vs General vs Specific)

@@

\subsection{Control Flow Analysis}

\subsubsection{context/path/flow-sensitive}

>>>>

\citeauthor{Boucher:1996:ACN:647473.727587} introduced abstract compilation as a new
implementation technique for abstract interpretation based static analysis \cite{Boucher:1996:ACN:647473.727587}.
The idea is inspired by partial evaluation and similar with the presented paper -- the program can be known 
at compile-time therefore the overhead if interpretation can be eliminated. The compiled analysis
can be represented by text or closures (higher-order functions).
Specifically, \citeauthor{Boucher:1996:ACN:647473.727587} show how to compile a monovariant control-flow 
analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007} 
for continuation-passing style (CPS) programs. Since the analyzed program is in CPS, the analyzer 
is essentially a big-step CE (\textit{Control-Environment}) abstract interpreter.
After specializing the analyzer with respect to the input program, the \textit{control} part of 
abstract interpreter is eliminated, instead, the compiled analysis is a closure which takes an 
environment as argument. Therefore the overhead of traversing the abstract syntax tree of input 
program also has been eliminated.

In this section, we show that \citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation can be 
understood and implemented as an instance of staging abstract interpreters.
We firstly revisit abstract compilation in the style of closure generation,
and then reproduce their result by adding stage annotations to the unstaged 0-CFA algorithm.
The generated program of our approach improves approximately the same extent of speed (roughly a factor of 2),
but without changing a single line of the analyzer program (with the use of LMS), 
while closure generation requires more engineering effort, such as a whole-program 
conversion on the analyzer. Moreover, our approach is able to not only remove the interpretive
overhead, but also specialize the data structures used in the analysis, for example, the store
that maps variables to sets of lambda.

\subsubsection{Closure Generation}

\subsubsection{Staged 0-CFA}

\subsubsection{Specializing the Data Structure}

@@

\subsection{Control-flow Analysis}

The target language we presented in section \ref{bg_lang} is essentially a higher-order functional language.
One fundamental analysis task for functional programs is the control-flow analysis, i,e., determining
which functions will possibly be applied at each call-site. 
Though, in the previous section, we discussed abstraction compilation and staging on a monovariant 
control-flow analysis for programs written in continuation-passing style; 
in this section, based on the staged abstract interpreter we presented in section \ref{sai}, 
we further discuss staging control-flow analysis for direct style languages in depth, 
especially polyvariant analyses.

\subsubsection{Context/Path/Flow-Sensitivity}
